A Type-Theoretical Wide-Coverage Computational Grammar for Swedish

### 2012-03-20 tsd/TSD2012/paper.tex
tsd paper, evaluation

% Chapter 2.4(with emphasis on the syntactic structures)
\textbf{Too long?}
Swedish is a North-Germanic language, sharing most of its grammatical structure 
with Norwegian and Danish. It is also one of the official languages in Finland
and is spoken by approximately 9 million people.
Although Swedish syntax often is similar to English, the  morphology is richer and the
word order more intricate.
It is a verb-second language \cite[p.116]{gunlog}: the
second constituent of a declarative main clause must consist of a verb.
The normal word order is subject-verb-object, but fronting other constituents
-- topicalisation -- is very common,  especially for temporal and
locative adverbial phrases \cite[\textsection 1027]{H&H}.
Fronting the finite verb marks questions.

@@

%Everything about Saldo, Lexin, etc, compressed.
%
%We use the resources from the other paper, so to say :)
\textbf{Some more here}
Our main lexical resource is SALDO \cite{}. We use the tools
described in \cite{} to extract a large-scale lexicon and annotate
this with valencies from Lexin. This gives us a dictionary with more than 
100 000 entries, covering all but 500 words from Talbanken ignoring names,
numbers and compounds.

@@

%Chapter 5 - short introduction about Talbanken tags + difference in
%the tags + 5.2(the translation process - first part) + example of
%2 trees  
Why we do this...... when is meta used
Talbanken05 uses three kinds of tags: categories, edge labels and POS-tags. 
While the POS-tags are reserved for words, the categories give grammatical information,
such as \verb|S|, \verb|NP| and \verb|VP|.
The edge labels give the
part of sentence: \verb|SS| for subject, 
\verb|OO| for object etc. Nearly 400 different tags are used, of which more than
300 different are POS-tags. The high number is due to the level of detail,  there are for
example almost 50 tags for nouns, excluding proper names. The tags
show definiteness, case and whether the word is a compound. Some words also have
their own tags.\\

@@

The evaluation measures the numbers of \textit{metas} in each translated tree.
If the finite verb is untranslatable, we get metas for the verb itself and for the
tense.
For all of Talbanken, we get a number of 65 \%, and if we limit our input to
uncomplicated sentences with no more than 10 words where all words are known to
our lexicon, we can reach 85 \%. When translating the whole treebank, 78 sentences
could not be mapped at all and resulted in \textit{metas} only.

\section{Chunk Parsing}

%New section - insights about latest work
The grammar, which is hand written, cannot be expected to cover all
possible expressions of the language.
Hence, we combine it with statistical means to get good coverage. Our approach
makes use of the rich annotation in Talbanken -- we perform chunk parsing
relying on some of the tags given in the treebank. 
Unknown words and grammatical constructions, idioms or ellipses cannot be
handled from within the grammar.
Therefore, whenever a sentence is not recognized by the grammar, it is 
parsed chunk by chunk.
Noun phrases, adverbial phrases and ? are considered. At the first stage,
we try to parse the basic sentence structure, and therefore allow the parser
to parse complicated chunks as dummy words. This way, we don't loose the 
high-level sentence analysis given by GF.
When this is done, we try to separately parse the chunks,
until as much of the sentence as possible is covered.

This way we get good coverage -- from ca 10 \% of Talbanken -- to ??

@@

 Maybe move this to future work, we have it but not really sure
 if it is actually used atm.

@@

This model is used for disambiguation: the parse trees for each sentence
are ranked by the sum of the probability of each function used.

@@

Section 5.3 + the evaluation part from 6.1

Latest results from chunk parsing and disambiguation

@@

abbrivations, idioms, learning adVs.
\textbf{too much about grammar details? there should be other things here as well}
At the end of the current project, we are left with 
many interesting future directions.

@@

\begin{itemize}
\item
\textit{Pronominal object shift} is common in Swedish and obligatory in Danish and Norwegian.
\enumsentence{\begin{tabular}[t]{@{}*{11}{l@{\ }}}
a. & Jag & ser&  honom& \textbf{inte}&~~ & b.&Jag &ser &\textbf{inte} &honom\\
   & \emph{I} & \emph{see}&\emph{him}&\emph{not}&&& \emph{I} &\emph{see}&\emph{not}&\emph{him}\\
\end{tabular}}
Personal pronouns are allowed to precede negation and some other sentence adverbials % --\textit{satsadverbial} --
in main clauses without auxiliary verbs.

\enumsentence{\begin{tabular}[t]{@{}*{12}{l@{\ }}}
a. & *Vi & har&  honom&  inte&sett&~~ & b.& * Jag &ser &huset &inte\\
   & \emph{~we} & \emph{have}&  \emph{him}&  \emph{not} &  \emph{seen} && b.& \emph{~I}&\emph{see} &\emph{the house} &\emph{not}\\
\end{tabular}}

Although object shifts are frequently used, they are hardly found in
Talbanken's part P, which has been the inspiration for this project.
Therefore, this implementation has so far not been prioritized.\\

\item
\textit{Bare indefinites} are at this point always parsed as mass nouns
    \enumsentence{\shortexnt{3}
{Jag & dricker& vatten}
{\emph{I}&\emph{drink}&\emph{water}}
}\label{sent:water}
although this clearly is a unsatisfying analysis for bare indefinites as in (\ref{sent:hund})
\enumsentence{\shortex{4}
{Vi & ska& köpa & hund}
{\emph{we} &\emph{will}& \emph{buy} & \emph{dog} }
{``We are going to buy a dog"}}\label{sent:hund}
\vspace{3mm}

\item
\textit{Object and subject control verbs}
The implementation of reflexive pronouns can be improved.  
It should for example be possible to differentiate between 
object and subject control verbs.

\eenumsentence{
\item[a]\shortex{7}
{Han & ber & henne & att & städa & sitt & rum}
{\emph{he} & \emph{begs} & \emph{her} & \emph{to} & \emph{clean} & {\sc self's} & \emph{room}}
{``He asks her to clean her room"}
\item[b]\shortex{7}
{Han& lovar& henne& att& städa &sitt & rum}
{\emph{he}& \emph{promises}& \emph{her}& \emph{to}&  \emph{clean} &\sc self's&\emph{room}}
{``He promises her to clean his room"}}
\end{itemize}

@@

@@

The grammar has been extended and enhanced, and its current status is
a specialized extension of the resource grammar.
Besides parsing, the grammar may well be used for language generation,
which works fast even when using an extensive lexicon.
Although it is not been formally verified, we believe that the majority of the
sentences generated are grammatically correct in a syntactical point of view.
Without any semantic knowledge, nonsense phrases cannot be avoided in
random generation. However, given that the abstract tree has a meaningful
interpretation,
the linearization should be correct. \\

Here we have some nice new results... (todo..)

\bibliographystyle{splncs}
\bibliography{FinBib}

>>>>

% Chapter 2.4(with emphasis on the syntactic structures)
\textbf{Too long?}
Swedish is a North-Germanic language, sharing most of its grammatical structure 
with Norwegian and Danish. It is also one of the official languages in Finland
and is spoken by approximately 9 million people.
Although Swedish syntax often is similar to English, the  morphology is richer and the
word order more intricate.
It is a verb-second language: the
second constituent of a declarative main clause must consist of a verb.
The normal word order is subject-verb-object, but fronting other constituents
-- topicalisation -- is very common,  especially for temporal and
locative adverbial phrases \cite[\textsection 1027]{H&H}.
Fronting the finite verb marks questions.

@@

%Everything about Saldo, Lexin, etc, compressed.
%
%We use the resources from the other paper, so to say :)
\textbf{Some more here}
Our main lexical resource is SALDO \cite{saldo}. We use the tools
described in \cite{ahlbergEnache} to extract a large-scale lexicon and annotate
this with valencies from Lexin. This gives us a dictionary with more than 
100 000 entries, covering all but 500 words from Talbanken ignoring names,
numbers and compounds.

@@

%Chapter 5 - short introduction about Talbanken tags + difference in
%the tags + 5.2(the translation process - first part) + example of
%2 trees  
Talbanken contains much valuable information about phrase structures and word 
usage. One part of this project has focused on translating trees from Talbanken to GF
by constructing a mapping, which automatically transforms trees in the
Talbanken format to GF abstract syntax trees.
We hence get a comparison between the two annotations and at the same time
we get methods for extracting and translating to GF notation.

Talbanken05 uses three kinds of tags: categories, edge labels and POS-tags. 
While the POS-tags are reserved for words, the categories give grammatical information,
such as \verb|S|, \verb|NP| and \verb|VP|.
The edge labels give the
part of sentence: \verb|SS| for subject, 
\verb|OO| for object etc. Nearly 400 different tags are used, of which more than
300 different are POS-tags. The high number is due to the level of detail,  there are for
example almost 50 tags for nouns, excluding proper names. The tags
show definiteness, case and whether the word is a compound. Some words also have
their own tags.\\

@@

If a node fails to be translated, a \textit{meta variable} is in its place. This is also the
case when two sister nodes can be translated, but no rule is applicable for joining them
into one tree.

\section{Chunk Parsing}
\label{sec:chunk}
%New section - insights about latest work
The grammar, which is hand written, cannot be expected to cover all
possible expressions of the language.
Hence, we combine it with statistical means to get good coverage. 
The parser uses named entity recognition and a compounding analysis.
Further we need to be able to give output
even when encountering unknown words and grammatical constructions 
as well as idioms or ellipses.
Our approach makes use of the rich annotation in Talbanken -- we perform chunk parsing
relying on some of the tags given in the treebank. 
Therefore, whenever a sentence is not recognized by the grammar, it is 
parsed chunk by chunk.
Noun phrases, adverbial phrases and adjectival phrases are considered. Due to 
differences in the annotation between Talbanken and GF, verbs are not considered
as parsable chunks. The verbs need their particles and preposition to be parsed
correctly, and these are usually not in the same Talbanken chunk. Therefore,
we focus on whole sentences and noun phrases. At the first stage,
we try to parse the basic sentence structure, and therefore allow the parser
to parse complicated chunks as dummy words. This way, we don't loose the 
high-level sentence analysis given by GF.
When this is done, we try to separately parse the chunks,
until as much of the sentence as possible is covered.

%This way we get good coverage -- from ca 10 \% of Talbanken -- to ??

@@

@@

The model is used for disambiguation: the parse trees for each sentence
are ranked by the sum of the probability of each function used.
Both whole sentence and chunks -- phrases -- can be ranked using this
principle.

@@

For the extraction of a GF Treebank from Talbanken, (section \ref{sec:mapping})
the evaluation measures the numbers of \textit{metas} in each translated tree.
If the finite verb is untranslatable, we get metas for the verb itself and for the
tense.
For all of Talbanken, we get a number of 65 \%, and if we limit our input to
uncomplicated sentences with no more than 10 words where all words are known to
our lexicon, we can reach 85 \%. When translating the whole treebank, 78 sentences
could not be mapped at all and resulted in \textit{metas} only.

A mapping between GF and the Wall Street Journal of Penn Treebank has earlier been conducted \cite{gfMech}.
The percentage of restored nodes from Peen Treebank is higher than our results.
The reason for this may be the fact that English is syntactically less complicated
than Swedish.
Furthermore, the text in Talbanken are from various brochures, newspapers and text books,
where idiomatic expressions are more likely the be appear and the language
presumably less strict than in Wall Street Journal\footnote{www.wsj.com}.
Also, the Penn Treebank contains a lower number of tags, 82 compared
to more than 300 in Talbanken. With more tags, we get more information, but as the number 
increase, so does the amount of work of finding the correct translation 
for each combination of tags and writing rules that cover all constructions. \\

The evaluation of the parser is performed by testing it on Talbanken sentences
and is still being done. So far, our evaluation results cover 643 sentences -- about 10 \%
of the treebank. Without the robustness given by the chunk parser,
we can only parse 17.9 \% of the sentences. For noun phrases, we cover 76 \%.
By using the chunk parsing described in section
\ref{sec:chunk}, we identify the structure on 65 \% of the sentences. That
is, the parser can identify the verbs and how they relate to the noun phrases,
prepositions and particles. The noun phrases and adverbs, however, may not be
completely parsed. If a verb is unknown to the grammar, or if it is used with
another valency, other prepositions or particles than the lexicon has assigned
it, the sentence structure can not be identified. In that case
the parser returns chunks, showing the parse trees of identifiable sub-phrases.

%Section 5.3 + the evaluation part from 6.1 
%
%Latest results from chunk parsing and disambiguation

@@

At the end of the current project, we are left with 
many interesting future directions. 
We should make sure to make as much use of resources as possible -- Saldo
and Lexin provides more information than we so far have been able to extract.
For example, data about abbreviations and idioms should
be added to the parser. This could be extracted from Saldo as well as Talbanken.
As the valency information is so important to the GF grammar, the lexicon should
be annotated with more of this in order to get a good parser.

@@

\begin{itemize}
\item
\textit{Pronominal object shift} is common in Swedish and obligatory in Danish and Norwegian.
\enumsentence{\begin{tabular}[t]{@{}*{11}{l@{\ }}}
a. & Jag & ser&  honom& \textbf{inte}&~~ & b.&Jag &ser &\textbf{inte} &honom\\
   & \emph{I} & \emph{see}&\emph{him}&\emph{not}&&& \emph{I} &\emph{see}&\emph{not}&\emph{him}\\
\end{tabular}}
%Personal pronouns are allowed to precede negation and some other sentence adverbials % --\textit{satsadverbial} --
%in main clauses without auxiliary verbs.
%
%\enumsentence{\begin{tabular}[t]{@{}*{12}{l@{\ }}}
%a. & *Vi & har&  honom&  inte&sett&~~ & b.& * Jag &ser &huset &inte\\
%   & \emph{~we} & \emph{have}&  \emph{him}&  \emph{not} &  \emph{seen} && b.& \emph{~I}&\emph{see} &\emph{the house} &\emph{not}\\
%\end{tabular}}
%
%Although object shifts are frequently used, they are hardly found in
%Talbanken's part P, which has been the inspiration for this project.
%Therefore, this implementation has so far not been prioritized.\\

\item
\textit{Bare indefinites} 
%are at this point always parsed as mass nouns
%    \enumsentence{\shortexnt{3}
%{Jag & dricker& vatten}
%{\emph{I}&\emph{drink}&\emph{water}}
%}\label{sent:water}
%although this clearly is a unsatisfying analysis for bare indefinites as in (\ref{sent:hund})
\enumsentence{\shortex{4}
{Vi & ska& köpa & hund}
{\emph{we} &\emph{will}& \emph{buy} & \emph{dog} }
{``We are going to buy a dog"}}\label{sent:hund}
\vspace{3mm}

\item
\textit{Object and subject control verbs}
%It should for example be possible to differentiate between 
%object and subject control verbs.
%
\eenumsentence{
\item[a]\shortex{7}
{Han & ber & henne & att & städa & sitt & rum}
{\emph{he} & \emph{begs} & \emph{her} & \emph{to} & \emph{clean} & {\sc self's} & \emph{room}}
{``He asks her to clean her room"}
\item[b]\shortex{7}
{Han& lovar& henne& att& städa &sitt & rum}
{\emph{he}& \emph{promises}& \emph{her}& \emph{to}&  \emph{clean} &\sc self's&\emph{room}}
{``He promises her to clean his room"}}
\end{itemize}

@@

A thorougly evaluation is still to be done, this should be carried out by
measuring the agreement between all parsed chunks and their respective
translation from Talbanken. The evaluation should also be done manually.
For this, we rely on an expert in Swedish.

@@

The grammar has been extended and enhanced, and its current status is
a specialized extension of the resource grammar.
Besides parsing, the grammar may well be used for language generation,
which works fast even when using an extensive lexicon.

Open source, transparent, modular, explores the strengths and weakneses
of GF.
\bibliographystyle{splncs}
\bibliography{FinBib}
